---
title: Cox models for survival analysis
author: Mathilde Donnenfeld
date: 2018-10-31
output:
  revealjs::revealjs_presentation:
    fig_width: 8
    fig_height: 6
    fig_caption: true
    transition: none
    css: theme.css
    highlight: pygments
    reveal_options:
      slideNumber: true
---

<h1 class="title">Introduction</h1>

<!-- Fonts go here to prevent blank slide -->
<style>
@import url('https://fonts.googleapis.com/css?family=Merriweather|Open+Sans');
</style>
<!-- End fonts -->

## Overview

What is survival analysis?
What is a Cox model?
The assumption of proportional hazards
Modelling
Conclusions
What about real life?

---

<h1 class="title">What is survival analysis?</h1>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE
	, message = FALSE
	, warning = FALSE
	# for testing
	# , cache=TRUE
)
```

```{r}
# uncomment these if you need to install anything
# install.packages("tidyverse")   # for tidy data
# install.packages("revealjs")    # for building presentation
# install.packages("survival")   # for survival analysis
# install.packages("ggfortify")   # for plotting survival curves
```


```{r}
library(survival)       # survival analysis
library(ggfortify)      # plot survival curves
library(tidyverse)      # tidy everything
library(glue)           # for string formatting
theme_set(theme_bw())   # nice theme for plots
```
---

### What is survival analysis?

* Trying to analyse the survival time of a set of observations

* For this, we need to define:
  * Survival (a 'death' event, e.g. actual death, medical condition declared, churn of customer...)
  * Time to Event (or TTE, i.e. the time for which we observed the individual)

---

### What is survival analysis?

Please meet Gertrude.
[INSERT PHOTO OF GERTRUDE HERE]

Gertrude is a *Pachira Aquatica*

---

### What is survival analysis?

* Hidden idea: all observations will end up 'dying'

* Either we observe this 'death', in which case we measure the time between the beginning of the observation period and the 'death'

* Or we don't, in which case we measure the time between the beginning of the observation period and today. Those observations are called 'right-censored'.

---

### Quick look at our dataset

```{r, include=TRUE}
head(df, 10)
```



---

### Quick look at our dataset

```{r}
summary(df$died)
summary(df$days)
```

---

### Define a survival object

```{r}

# Using the Surv function:

df$survival <- Surv(time=df$days, event=df$died)


```

---
###Plot the survival curve
```{r}
fit_proportional <- survfit(survival ~ species, data = df)
autoplot(fit_proportional)
```

---

```{r}

survdiff(survival ~ species,data=df)

```
---

### What is a Cox model?

* Regression with the hazard function as dependant variable

* h(t) = h0(t) * exp(ß1 x1 +  ß2 x2 + ... + ßn xn)

* Only the baseline hazard function varies with time

---

### The assumption of proportional hazards

```{r}
fit <- survfit(survival ~ name, data = df)
autoplot(fit)
```

---
### Modelling

In code:

```{r, include=TRUE}

model_full <- coxph(survival ~ name
      + species
, data = df)
```

---  

### Testing assumption of proportional hazards

```{r}
cox.zph(model_full)
```

---

### Modelling: simplification of the model

```{r}

model_short <- coxph(survival ~ species
, data = df)

cox.zph(model_short)

```

---

### Modelling: Model output

```{r}
summary(model_short)
```

---
### Takeaways

* Survival analysis is a powerful tool for analyzing the association of variables with a hazard function

* With great power comes a great assumption

* Congrats to Gertrude (1.5 years and counting!)

---
### What about real life?

In real life, several issues can come across:

* The proportional hazards assumption is a strong one, and hard to meet.

* This is especially true for large datasets

* It can make sense to randomly select a subsample of your population

---



## Further reading
* Walters, *What is a Cox model?*
    * [link](http://www.bandolier.org.uk/painres/download/whatis/COX_MODEL.pdf)
* CRAN, *Plotting with survival package*
    * [link](https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_surv.html)
* LaMorte, *Comparing survival curves: The log-rank test*
    * [link](http://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/BS704_Survival/BS704_Survival5.html)


---

<h1 class="title">Thanks!</h1>